---
title: "Week 3 Probability Assignment"
author: "Benjamin Siebold"
date: "`r format(Sys.time(), '%b, %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\definecolor{hightlightColor}{HTML}{FFFF66}

## Introduction

### Question 1

1. In a study of pleas and prison sentences, it is reported that 42% of the subjects were sent to prison. Among those sent to prison, 38% plead guilty. Among those not sent to prison, 50% plead guilty.

With the above information, the following probability assignments are made:  
P(P) = sent to prison  
P(!P) = not sent to prison  
P(G|P) = plead guilty prison   
P(G|!P) = plead guilty no prison   
P(!G|P) = plead not guilty prison   
P(!G|!P) = plead not guilty no prison   

a. If a subject is randomly selected, what is the probability of getting a person who was not sent to prison.  
P(!P) = 1 - P(P)

```{r P1, include=TRUE}
1-.42
```
\colorbox{hightlightColor}{P(!P) = .58}  

b. If a subject is randomly selected, and it is known that the subject entered a guilty plea, what is the probability that this subject was not sent to prison.

unknown: P(!P|G)  
known:  
P(P) = .42  
P(!P) = .58  
P(G|P) = .38  
P(!G|P) = .62  
P(G|!P) = .50  
P(!G|!P) = .50  

equation: P(!P|G) = P(G|!P)P(!P)/(P(G|!P)P(!P)+P(G|P)P(P))
      
```{r P2, include=TRUE}
(.58)*(.50)/((.58)*(.50)+(.42)*(.38))
```
\colorbox{hightlightColor}{P(~P|G) = .645}  

c. If a subject is randomly selected, what is the probability of getting someone who was sent to prison.

\colorbox{hightlightColor}{The answer is given in the description of the problem: .42}  

d. If a subject is randomly selected, and it is known that the subject entered a guilty plea, what is the probability that this person was sent to prison.

unknown: P(P|G)  
known:   
P(P) = .42    
P(!P) = .58  
P(G|P) = .38   
P(!G|P) = .62  
P(G|!P) = .50  
P(!G|!P) = .50  

equation: P(P|G) = P(G|P)P(P)/(P(G|P)P(P)+P(G|!P)P(!P))
```{r P4, include=TRUE}
(.38)*(.42)/((.38)*(.42)+(.50)*(.58))
```
\colorbox{hightlightColor}{P(P|G) = .355}  

### Question 2

Given the following table:  

\begin{figure}[h!]
\includegraphics[width=0.75\textwidth]{customer_table.PNG}
\end{figure}

a. What is the value of each of the following probability? 

The below can be determined by looking at the number of rows in which both conditions are true. For example in p1 there is a probability that Gender = M of 10/20, but only 6 of those rows have class = C0, thus, 6/20.  
\colorbox{hightlightColor}{P(Gender = M | Class = C0) = 6/20}  
\colorbox{hightlightColor}{P(Gender = F | Class = C1)  = 6/20}    
\colorbox{hightlightColor}{P(Car Type = Family | Class = C0) = 1/20}     
\colorbox{hightlightColor}{P(Car Type = Family | Class = C1) = 3/20}  
\colorbox{hightlightColor}{P(Shirt Size = Medium | Class = C0) = 3/20}  
\colorbox{hightlightColor}{P(Shirt Size = Medium | Class = C1) = 4/20}  

b. Use Naive Bayes' Classifier to find the class of:

P(Gender=F, Car Type = Family, Shirt Size = Medium)

Class=C0  

P(X|Class=C0) = P(Gender=F|Class=C0) *  
P(Car Type=Family|Class=C0) *  
P(Shirt Size = Medium|Class=C0)
```{r P(X|C0), include=TRUE}
(4/20)*(1/20)*(3/20)
```
Class=C1  

P(X|Class=C1) = P(Gender=F|Class=C1) *  
P(Car Type=Family|Class=C1) *  
P(Shirt Size = Medium|Class=C1)
```{r P(X|C1), include=TRUE}
(6/20)*(3/20)*(4/20)
```

\colorbox{hightlightColor}{Because P(X|Class=C1) > P(X|Class=C0) (0.009 > 0.0015)}  
\colorbox{hightlightColor}{the class of P(Gender=F, Car Type = Family, Shirt Size = Medium) = C1}
